\documentclass{article}

\usepackage{geometry}
\geometry{a4paper}

\usepackage[round,authoryear]{natbib}
\bibliographystyle{plainnat}
\usepackage[hidelinks]{hyperref}
\usepackage{authblk}
\usepackage{tabularray}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{xcolor}

% Make affiliations run inline
\renewcommand\Affilfont{\small}
\renewcommand\Authfont{\normalsize}

\makeatletter
\renewcommand\AB@affilsepx{;\hspace{0.6em}\protect\Affilfont}
\makeatother

\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesubsection}{\arabic{subsection}}

\begin{document}

\title{Population-scale Ancestral Recombination Graphs with tskit 1.0\\
\large Supplementary Information}

\input{authors.tex}

\date{}
\maketitle

\subsection*{Published tools using tskit}
We performed a search for published software using tskit.
To be included in this list, the software must be described in a publication
as a named tool intended for reuse and must use tskit.
This list is intentionally conservative, and does not include
(for example) publications depending on bespoke analysis
pipelines using tskit.
For each tool in Table S1, we list: its name (with a clickable
hyperlink to its source code repository);
the main implementation
language(s) of the tool based on GitHub summaries;
the DOI (with clickable hyperlink) of the publication in which this
tool was first described (or, in the case of SLiM, the first
publication in which tskit-based features were described);
and the year of publication and number of citations to this
publication according to data from OpenAlex
(\url{https://openalex.org}) retrieved on 2026-01-23.

\input{tools_table.tex}

\subsection*{Tskit functionality and benchmarks}


Tskit is the result of more than a decade of sustained development,
and its distinctive operations have been introduced in a
series of publications. The succinct tree
sequence data format and the sequential tree-construction algorithm
were first introduced as part of the msprime coalescent
simulator~\citep{kelleher2016efficient}.
The key algorithmic result is
that, given the edge table of an ARG, the cost of iterating over
the marginal trees along the genome scales with the number of trees
and the edge differences between neighbouring trees, rather than
with the sample size.
The practical consequence was demonstrated in a benchmark
where using the succinct tree sequence format and sequential
iteration was more than a million times faster than reading the
same trees individually from Newick using the standard libraries
available at the time, for a simulation of 100,000 samples, with
storage requirements reduced by roughly four orders of magnitude.
The data model was subsequently generalised to enable
forwards-time simulation~\citep{kelleher2018efficient}.
This publication
also introduced the columnar tables API and was the first paper in
which the tskit library itself appeared as a distinct component.
That paper introduced the ``simplify'' operation, which restricts an ARG to the
history ancestral to a specified set of samples, showing that its
time complexity is linear in the number of input edges.
ARG simplification is a fundamental operation with many
applications~\citep{wong2024general},
as well as being the primitive operation making forwards-time simulation of ARGs
tractable~\citep{haller2018tree}.
The pyslim library~\citep{gopalan2025bridging} bridges SLiM and
tskit at the Python level, operationalising these hybrid
workflows together using tskit's ``union'' operation for
stitching together tree sequences from independent simulations;
this combination has enabled whole-chromosome forward simulations
of the entire 10-million-year history of the great
apes~\citep{rodrigues2024shared} and
whole-genome human simulations under realistic demographic and selection
models~\citep{haller2025simHumanity}.

Following this foundational work, tskit
accumulated a sequence of new operations, each first described in its own
publication. Motivated by needing to summarise biobank-scale inferred ARGs
with uncalibrated node times, \citet{kelleher2019inferring} introduced
the genealogical nearest neighbours (GNN) statistic, and
demonstrated that the tree-sequence encoding compresses inferred
biobank-scale data by orders of magnitude relative to VCF.
\citet{ralph2020efficiently} then introduced a general API for
population-genetic statistics built around a duality theorem
relating site-based and branch-based statistics.
The benchmarks in that paper show
speedups of roughly three orders of magnitude, together with
a $\sim 200$-fold reduction in memory footprint, against
genotype-matrix implementations at a sample size of one million haplotypes.
\citet{wohns2022unified} subsequently introduced the tsdate method
for estimating ages of ancestral nodes from an inferred ARG under
a mutational-clock prior, and used it to combine 3{,}601 modern
and eight high-coverage ancient human genomes into a single
unified genealogy.
The msprime~1.0 release~\citep{baumdicker2022efficient} provided further
illustrations of the flexibility of the data model,
including simulations of gene conversion and complete ARGs.
Individual and pedigree abstractions were added to the tskit data
model by \citet{andersontrocme2023genes}, extending what a node or
individual in an ARG can represent.
\citet{tsambos2023link} introduced the ``link\_ancestors''
operation, a new primitive for extracting local-ancestry tracts
from an ARG.

Development of the core API has continued through
2024--2026. \citet{fritze2026forest} added the
``extend\_haplotypes'' function to tskit, which restores
unary-node spans lost to simplification or inference and roughly
halves the number of edges in typical ARGs, with corresponding
reductions in statistic-computation time.
\citet{lehmann2026on}
introduced the operation of computing genetic relatedness matrices
(GRMs) directly from an ARG, expressing GRM--vector products as
tree traversals: the branch GRM--vector product for
one million diploids on human chromosome~21 completes in
under 18~seconds, and the associated randomised PCA scales to
$2^{20}$ samples where genotype-matrix PCA runs out of memory.
\citet{lee2026genetic} apply this
last operation to fitting linear mixed models and provide the one recent
head-to-head benchmark of tskit against another library,
comparing the ARG-based GRM with the corresponding implementation
in ARGneedle-lib~\citep{zhang2023biobank},
and reporting that tskit is one to two orders of
magnitude faster for GRM--vector products across a wide range of
mutation rates,
while using an exact algorithm that avoids Monte-Carlo approximation.
Alongside these core additions, the exposure of tskit's
traversal orders to Numba~\citep{lam2015numba} has allowed
Python-level user code to JIT-compile ARG traversals at close to C
speed, enabling downstream packages such as
tstrait~\citep{tagami2024tstrait} and tsbrowse~\citep{karthikeyan2025tsbrowse}
to perform biobank-scale computations in pure Python.
Real-data applications have further stressed the data model
outside its original coalescent-simulation setting: an ARG for
2.48~million SARS-CoV-2 genomes inferred with sc2ts occupies only
32\,MiB and loads into memory in about half a
second~\citep{zhan2025pandemic}; and recent work has dated
contiguous ARGs spanning 14\% of the human genome for 47,535
Genomics England participants, estimating ages for 23.2 million
variants~\citep{pope2026tracing}.

Tskit~1.0 is the point at which this
decade-long accumulation of operations is brought together and
stabilised under a single, guaranteed API.
Table S2 lists properties and operations it supports, grouped
into nine broad categories.
We contrast tskit's functionality with three
Python libraries from adjacent niches:
ARGneedle-lib~\citep{zhang2023biobank}, which infers and analyses
ancestral recombination graphs from biobank-scale data;
the BTE interface (\url{https://github.com/jmcbroome/BTE})
for matUtils~\citep{mcbroome2021daily},
a toolkit for manipulating and analysing mutation-annotated trees
of SARS-CoV-2; and DendroPy~\citep{sukumaran2010dendropy}, a
mature and popular phylogenetic scripting library.
For each operation we mark whether the comparison library offers a
fully comparable feature:
\checkmark{} denotes full
support, $\circ$ denotes partial or restricted support, and
a blank cell denotes none.
Each refers to documented methods in the Python library;
additional functionality may be available in the underlying code
(for instance, ARGneedle-lib's C++ code can be used to do
efficient tree traversal and iteration, and this underlies
other functionality in ARGneedle-lib).
The comparisons were performed with
tskit version 1.0.2,
arg-needle-lib version 1.2.1,
BTE version 0.9.3,
and DendroPy version 5.0.8.

\input{functionality_table.tex}

\bibliography{paper}

\end{document}
